Spike-and-Slab Generalized Additive Models and Scalable Algorithms for High-Dimensional Data

### 2021-10-21 Manuscript/03-Simulation.Rmd
Revise Simulation

The results are tabulated in Table \ref{tab:gaus}, \ref{tab:bin_auc}.
\begin{center}
Table \ref{tab:gaus} here\\
Table \ref{tab:bin_auc} here\\
\end{center}
```{r, fig.height=8, eval = FALSE}
targets::tar_load(binom_plot)
binom_plot
```

@@

The predictive performances have a consistent pattern across the three distributions of outcomes. Across all the scenarios, COSSO and adaptive COSSO have the least favorable performance among the applicable methods examined. To note, mgcv doesn't support high-dimensional analysis when the number of coefficients are greater than the sample size. mgcv predicts well when $p$ is small or moderate (p = 4, 10), and deteriorate when the number of predictors increase. Among the three fast-computing Bayesian hierarchical models, the proposed models, BHAM-IWLS and BHAM-CD predicts better than SB-GAM when the dimension of are small, moderate, and high. However, in the hyper high dimensional case, SB-GAM has better performance. Among the proposed methods, BHAM-IWLS performs consistently better than BHAM-CD.

>>>>

\begin{center}
Table \ref{tab:gaus} here\\
Table \ref{tab:bin_auc} here\\
\end{center}
```{r, fig.height=8, eval = FALSE}
targets::tar_load(binom_plot)
binom_plot
```

@@

The predictive performances have a consistent pattern across the two distributions of outcomes. Across all the scenarios, COSSO and adaptive COSSO have the least favorable performance among the applicable methods examined (See Table \ref{tab:gaus} and \ref{tab:bin_auc}). To note, mgcv doesn't support high-dimensional analysis, i.e. the number of coefficients are greater than the sample size, and hence not evaluated when $p=100, 200$. mgcv predicts well when $p$ is small or moderate ($p = 4, 10$), and deteriorate when the number of predictors increase. Among the three fast-computing Bayesian hierarchical models, the proposed models, BHAM-IWLS and BHAM-CD predicts better than SB-GAM when the dimension of are moderate (p=4, 10, 50). Particularly, BHAM-IWLS performs as good as mgcv if not better. However, in high dimensional case where we mimic the situation the signals are extremely sparse, SB-GAM has better performance than the proposed method. However, the BHAM-CD has extreme computational advantage over SB-GAM (see Table \ref{tab:time_sim}) without sacrificing much of the prediction accuracy.

### 2021-10-21 Manuscript/04-Real_Data.Rmd
Revise Simulation

In this section, we apply the proposed models BHAM, fitted with the EM-CD algorithm, to analyze two real-world metabolomics datasets where the outcomes are binary and continuous respectively. We demonstrate the improved prediction performance compared to the other Bayesian hierarchical additive model, SB-GAM [@Bai2021]. Computation time for cross-validation and optimal model fitting steps are recorded and presented in Table \ref{tab:time_real_data}.
\begin{center}
Table \ref{tab:time_real_data} here
\end{center}

## Emory Cardiovascular Biobank
\label{sec:ECB}
We use the proposed models BHAM to analyze a metabolic dataset from a recently published research [@Mehta2020] studying plasma metabolomic profile on the three-year all-cause mortality among patients undergoing cardiac catheterization. The dataset is publicly available via \textit{Dryad} [@Mehta2020_data]. It contains in total of 776 subjects from two cohorts. As there is a large number of non-overlapping features among the two cohorts, we use the cohort with larger sample size (N=454). There are initially 6796 features in the dataset, which is too large to be practically meaningful to analyze. Hence, we perform a univariate screening procedure on the features, via GAM implemented in \texttt{mgcv}, and choose the the top 200 features with smallest p-values. We use 5-knot spline additive models for binary outcome using two different models, the proposed BHAM and the SB-GAM. 10-Fold CV are used to choose the optimal tuning parameters of each framework with respect to the default selection criterion implemented in the software. Out-of-bag samples are used for prediction performance evaluation, where deviance, AUC, Brier score defined as $\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2$, and misclassication error defined as $\frac{1}{n}\sum\limits^{n}_{i=1}I(|y_i - \hat y_i|>0.5)$ are calculated. BHAM-CD obtains superior AUC, Brier score, and misclassification error in the out-of-bag samples compared to SB-GAM (Table \ref{tab:ECB_res}).

@@

## Weight Loss Maintenance Cohort 
We use the proposed models BHAM to analyze metabolomics data from a recently published study [@Bihlmeyer2021] on the association between metabolic biomarkers and weight loss, where the dataset is publicly available [@Bihlmeyer2021_data]. In this analysis, we primarily focuse on the analysis of one of the three studies included, weight loss maintenance cohort [@Svetkey2008], due to the drastically different intervention effects. In the dataset, 765 metaboliltes in baseline plasma collected were profiled using liquid chromatography mass spectrometry. Quality control and natural log transformation are performed during metabolites data preparation. The outcome of interest are standardized percent change in insulin resistance, and hence modeled using a Gaussian model. After removing missing datapoints and addressing outliers in the data, there are $p$=`r nrow(WLM_train_dat)` features remaining in the analysis. 5-Knot spline additive models for the Gaussian outcome are constructed using two different models, the proposed BHAM and the SB-GAM. 10-Fold CV are used to choose the optimal tuning parameters of each framework with respect to the default selection criterion implemented in the software. Out-of-bag samples are used for prediction performance evaluation, where deviance, $R^2$, mean squared error (MSE) defined as $\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2$, and mean absolute error (MAE) defined as $\frac{1}{n}\sum\limits^{n}_{i=1}|y_i - \hat y_i|$ are calculated. BHAM-CD obtains superior $R^2$, MSE, and MAE in the out-of-bag samples compared to SB-GAM (Table \ref{tab:WLM_res}).
\begin{center}
Table \ref{tab:WLM_res} here
\end{center}

>>>>

In this section, we apply the proposed models BHAM, fitted with the EM-CD algorithm, to analyze two real-world metabolomics datasets where the outcomes are binary and continuous respectively. We demonstrate the improved prediction performance compared to the other Bayesian hierarchical additive model, SB-GAM [@Bai2021], while being computationally efficient (see Table \ref{tab:time_real_data}).
\begin{center}
Table \ref{tab:time_real_data} here
\end{center}

## Emory Cardiovascular Biobank
\label{sec:ECB}
We use the proposed models BHAM to analyze a metabolic dataset from a recently published research [@Mehta2020] studying plasma metabolomic profile on the three-year all-cause mortality among patients undergoing cardiac catheterization. The dataset is publicly available via \textit{Dryad} [@Mehta2020_data]. It contains in total of 776 subjects from two cohorts. As there is a large number of non-overlapping features among the two cohorts, we use the cohort with larger sample size (N=454). There are initially 6796 features in the dataset, which is too large to be practically meaningful to analyze. Hence, we perform a univariate screening procedure on the features, via GAM implemented in \texttt{mgcv}, and choose the the top 200 features with smallest p-values. We use 5-knot spline additive models for binary outcome using two different models, the proposed BHAM and the SB-GAM. 10-Fold CV are used to choose the optimal tuning parameters of each framework with respect to the default selection criterion implemented in the software. Out-of-bag samples are used for prediction performance evaluation, where deviance, AUC, Brier score defined as $\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2$, and misclassication error defined as $\frac{1}{n}\sum\limits^{n}_{i=1}I(|y_i - \hat y_i|>0.5)$ are calculated. BHAM-CD obtains superior AUC, Brier score, and misclassification error in the out-of-bag samples compared to SB-GAM (see Table \ref{tab:ECB_res}).

@@

## Weight Loss Maintenance Cohort 
We use the proposed models BHAM to analyze metabolomics data from a recently published study [@Bihlmeyer2021] on the association between metabolic biomarkers and weight loss, where the dataset is publicly available [@Bihlmeyer2021_data]. In this analysis, we primarily focuse on the analysis of one of the three studies included, weight loss maintenance cohort [@Svetkey2008], due to the drastically different intervention effects. In the dataset, 765 metaboliltes in baseline plasma collected were profiled using liquid chromatography mass spectrometry. Quality control and natural log transformation are performed during metabolites data preparation. The outcome of interest are standardized percent change in insulin resistance, and hence modeled using a Gaussian model. After removing missing datapoints and addressing outliers in the data, there are $p$=`r nrow(WLM_train_dat)` features remaining in the analysis. 5-Knot spline additive models for the Gaussian outcome are constructed using two different models, the proposed BHAM and the SB-GAM. 10-Fold CV are used to choose the optimal tuning parameters of each framework with respect to the default selection criterion implemented in the software. Out-of-bag samples are used for prediction performance evaluation, where deviance, $R^2$, mean squared error (MSE) defined as $\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2$, and mean absolute error (MAE) defined as $\frac{1}{n}\sum\limits^{n}_{i=1}|y_i - \hat y_i|$ are calculated. BHAM-CD obtains superior $R^2$, MSE, and MAE in the out-of-bag samples compared to SB-GAM (see Table \ref{tab:WLM_res}).
\begin{center}
Table \ref{tab:WLM_res} here
\end{center}

### 2021-10-21 Manuscript/SS_GAM.tex
Revise Simulation

The results are tabulated in Table \ref{tab:gaus}, \ref{tab:bin_auc}.

@@

The predictive performances have a consistent pattern across the three
distributions of outcomes. Across all the scenarios, COSSO and adaptive
COSSO have the least favorable performance among the applicable methods
examined. To note, mgcv doesn't support high-dimensional analysis when
the number of coefficients are greater than the sample size. mgcv
predicts well when \(p\) is small or moderate (p = 4, 10), and
deteriorate when the number of predictors increase. Among the three
fast-computing Bayesian hierarchical models, the proposed models,
BHAM-IWLS and BHAM-CD predicts better than SB-GAM when the dimension of
are small, moderate, and high. However, in the hyper high dimensional
case, SB-GAM has better performance. Among the proposed methods,
BHAM-IWLS performs consistently better than BHAM-CD.

@@

In this section, we apply the proposed models BHAM, fitted with the
EM-CD algorithm, to analyze two real-world metabolomics datasets where
the outcomes are binary and continuous respectively. We demonstrate the
improved prediction performance compared to the other Bayesian
hierarchical additive model, SB-GAM \citep{Bai2021}. Computation time
for cross-validation and optimal model fitting steps are recorded and
presented in Table \ref{tab:time_real_data}.

@@

\label{sec:ECB} We use the proposed models BHAM to analyze a metabolic
dataset from a recently published research \citep{Mehta2020} studying
plasma metabolomic profile on the three-year all-cause mortality among
patients undergoing cardiac catheterization. The dataset is publicly
available via \textit{Dryad} \citep{Mehta2020_data}. It contains in
total of 776 subjects from two cohorts. As there is a large number of
non-overlapping features among the two cohorts, we use the cohort with
larger sample size (N=454). There are initially 6796 features in the
dataset, which is too large to be practically meaningful to analyze.
Hence, we perform a univariate screening procedure on the features, via
GAM implemented in \texttt{mgcv}, and choose the the top 200 features
with smallest p-values. We use 5-knot spline additive models for binary
outcome using two different models, the proposed BHAM and the SB-GAM.
10-Fold CV are used to choose the optimal tuning parameters of each
framework with respect to the default selection criterion implemented in
the software. Out-of-bag samples are used for prediction performance
evaluation, where deviance, AUC, Brier score defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2\), and
misclassication error defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}I(|y_i - \hat y_i|>0.5)\) are
calculated. BHAM-CD obtains superior AUC, Brier score, and
misclassification error in the out-of-bag samples compared to SB-GAM
(Table \ref{tab:ECB_res}).

@@

We use the proposed models BHAM to analyze metabolomics data from a
recently published study \citep{Bihlmeyer2021} on the association
between metabolic biomarkers and weight loss, where the dataset is
publicly available \citep{Bihlmeyer2021_data}. In this analysis, we
primarily focuse on the analysis of one of the three studies included,
weight loss maintenance cohort \citep{Svetkey2008}, due to the
drastically different intervention effects. In the dataset, 765
metaboliltes in baseline plasma collected were profiled using liquid
chromatography mass spectrometry. Quality control and natural log
transformation are performed during metabolites data preparation. The
outcome of interest are standardized percent change in insulin
resistance, and hence modeled using a Gaussian model. After removing
missing datapoints and addressing outliers in the data, there are
\(p\)=237 features remaining in the analysis. 5-Knot spline additive
models for the Gaussian outcome are constructed using two different
models, the proposed BHAM and the SB-GAM. 10-Fold CV are used to choose
the optimal tuning parameters of each framework with respect to the
default selection criterion implemented in the software. Out-of-bag
samples are used for prediction performance evaluation, where deviance,
\(R^2\), mean squared error (MSE) defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2\), and mean
absolute error (MAE) defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}|y_i - \hat y_i|\) are calculated.
BHAM-CD obtains superior \(R^2\), MSE, and MAE in the out-of-bag samples
compared to SB-GAM (Table \ref{tab:WLM_res}).

>>>>

@@

The predictive performances have a consistent pattern across the two
distributions of outcomes. Across all the scenarios, COSSO and adaptive
COSSO have the least favorable performance among the applicable methods
examined (See Table \ref{tab:gaus} and \ref{tab:bin_auc}). To note, mgcv
doesn't support high-dimensional analysis, i.e.~the number of
coefficients are greater than the sample size, and hence not evaluated
when \(p=100, 200\). mgcv predicts well when \(p\) is small or moderate
(\(p = 4, 10\)), and deteriorate when the number of predictors increase.
Among the three fast-computing Bayesian hierarchical models, the
proposed models, BHAM-IWLS and BHAM-CD predicts better than SB-GAM when
the dimension of are moderate (p=4, 10, 50). Particularly, BHAM-IWLS
performs as good as mgcv if not better. However, in high dimensional
case where we mimic the situation the signals are extremely sparse,
SB-GAM has better performance than the proposed method. However, the
BHAM-CD has extreme computational advantage over SB-GAM (see Table
\ref{tab:time_sim}) without sacrificing much of the prediction accuracy.

@@

In this section, we apply the proposed models BHAM, fitted with the
EM-CD algorithm, to analyze two real-world metabolomics datasets where
the outcomes are binary and continuous respectively. We demonstrate the
improved prediction performance compared to the other Bayesian
hierarchical additive model, SB-GAM \citep{Bai2021}, while being
computationally efficient (see Table \ref{tab:time_real_data}).

@@

\label{sec:ECB} We use the proposed models BHAM to analyze a metabolic
dataset from a recently published research \citep{Mehta2020} studying
plasma metabolomic profile on the three-year all-cause mortality among
patients undergoing cardiac catheterization. The dataset is publicly
available via \textit{Dryad} \citep{Mehta2020_data}. It contains in
total of 776 subjects from two cohorts. As there is a large number of
non-overlapping features among the two cohorts, we use the cohort with
larger sample size (N=454). There are initially 6796 features in the
dataset, which is too large to be practically meaningful to analyze.
Hence, we perform a univariate screening procedure on the features, via
GAM implemented in \texttt{mgcv}, and choose the the top 200 features
with smallest p-values. We use 5-knot spline additive models for binary
outcome using two different models, the proposed BHAM and the SB-GAM.
10-Fold CV are used to choose the optimal tuning parameters of each
framework with respect to the default selection criterion implemented in
the software. Out-of-bag samples are used for prediction performance
evaluation, where deviance, AUC, Brier score defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2\), and
misclassication error defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}I(|y_i - \hat y_i|>0.5)\) are
calculated. BHAM-CD obtains superior AUC, Brier score, and
misclassification error in the out-of-bag samples compared to SB-GAM
(see Table \ref{tab:ECB_res}).

@@

We use the proposed models BHAM to analyze metabolomics data from a
recently published study \citep{Bihlmeyer2021} on the association
between metabolic biomarkers and weight loss, where the dataset is
publicly available \citep{Bihlmeyer2021_data}. In this analysis, we
primarily focuse on the analysis of one of the three studies included,
weight loss maintenance cohort \citep{Svetkey2008}, due to the
drastically different intervention effects. In the dataset, 765
metaboliltes in baseline plasma collected were profiled using liquid
chromatography mass spectrometry. Quality control and natural log
transformation are performed during metabolites data preparation. The
outcome of interest are standardized percent change in insulin
resistance, and hence modeled using a Gaussian model. After removing
missing datapoints and addressing outliers in the data, there are
\(p\)=237 features remaining in the analysis. 5-Knot spline additive
models for the Gaussian outcome are constructed using two different
models, the proposed BHAM and the SB-GAM. 10-Fold CV are used to choose
the optimal tuning parameters of each framework with respect to the
default selection criterion implemented in the software. Out-of-bag
samples are used for prediction performance evaluation, where deviance,
\(R^2\), mean squared error (MSE) defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}(y_i - \hat y_i)^2\), and mean
absolute error (MAE) defined as
\(\frac{1}{n}\sum\limits^{n}_{i=1}|y_i - \hat y_i|\) are calculated.
BHAM-CD obtains superior \(R^2\), MSE, and MAE in the out-of-bag samples
compared to SB-GAM (see Table \ref{tab:WLM_res}).
